---
title: "PDM SCM CNV analysis"
output: html_document
date: "2024-07-31"
---

```{r setup, include=FALSE}
library(tidyverse)
library(corrplot)
library(ggbiplot)
library(ggpubr)
library(cowplot)
library(ggdendro)
library(RColorBrewer)
library(RaggedExperiment)
library(factoextra)
library(GenomicRanges)
library(EnsDb.Hsapiens.v75)
source("Myfunctions.R")
set.seed(1234)
```

## Results {.tabset}  

### features associated with subtypes

Annotate segment by genes

```{r, message=FALSE, warning=FALSE}
# Genes in covered intervals
interval <- read.table("/vast/ai_projects/melanoma_depigmentation_biomarker_discovery/peinan/PDM_SCM/data/ref/resources/TST500C_manifest.bed", sep = "\t")
seq.genes <- unique(gsub("_.*","", interval$V4))

# Convert segment table to GRanges
PDM_seg <- read.table("/vast/ai_projects/melanoma_depigmentation_biomarker_discovery/peinan/PDM_SCM/data/GISTIC/PDM_segments.txt", sep = "\t", header = T) %>%
  dplyr::filter(Chromosome != "Y")
SCM_seg <- read.table("/vast/ai_projects/melanoma_depigmentation_biomarker_discovery/peinan/PDM_SCM/data/GISTIC/SCM_segments.txt", sep = "\t", header = T) %>%
  dplyr::filter(Chromosome != "Y")
segment_all <- rbind(PDM_seg, SCM_seg)
segment_grl.list <- makeGRangesListFromDataFrame(segment_all[,-c(5)], split.field="Sample", keep.extra.columns=TRUE)
segment_grl.list <- GenomicRanges::sort(segment_grl.list)

# Map to Ensembl gene annotation V75 GRCh37 (hg19)
hs.genes <- ensembldb::genes(EnsDb.Hsapiens.v75)
sel.genes <- subset(hs.genes, gene_name %in% seq.genes)
CNV.regions <- as(segment_grl.list, "RaggedExperiment")
qassay <- GrangetoQassay(CNV.regions, sel.genes, .largest)
data.CNV <- t(qassay) %>%
  as.data.frame()

names(data.CNV) <- paste(names(data.CNV), "_log2R", sep = "")
data.CNV <- data.CNV %>%
  mutate(Subtype = ifelse(grepl("PDM", rownames(data.CNV)), 0, 1), .before = 1)

# annotate genes by chromosomes
sel.genes <- sel.genes%>%
  as.data.frame() #%>%
  #mutate(gene_name = gsub("\\-", "\\.", gene_name))

gene.annotate <- data.frame(feature = grep("log2R", names(data.CNV), value = T)) %>%
  mutate(gene_name = gsub("_log2R", "", feature)) %>%
  left_join(dplyr::select(sel.genes, c("gene_name", "seqnames")), by = "gene_name") %>%
  dplyr::filter(seqnames %in% c(1:22, "X")) %>%
  mutate(gene_annotate = paste(gene_name, "_chr", seqnames, sep = "")) 

colnames(data.CNV)[match(gene.annotate$feature, names(data.CNV))] <- gene.annotate$gene_annotate
```  


```{r}
# H2O.CNV <- data.CNV %>%
#   mutate(sample = rownames(data.CNV), .before = 1,
#          Subtype = ifelse(Subtype == 0, "PDM", "metastasis")) %>%
#   left_join(dplyr::select(PureCN_summary, c("sample","purity", "ploidy","PGA_diploid","PGA_dominant","LOH_fraction")), by = "sample") %>%
#   left_join(filtered_mutations, by = c("sample" = "Tumor_Sample_Barcode")) %>%
#   dplyr::select(-sample)
# write.csv(H2O.CNV, "tables/PDM_all_features_H2O_input.csv", row.names = F)
```


Filter non-altered genes. 

```{r}
PDM.CNV <- data.CNV %>%
  dplyr::filter(grepl("PDM", rownames(data.CNV)))

SCM.CNV <- data.CNV %>%
  dplyr::filter(grepl("SCM", rownames(data.CNV)))

CNV.event.PDM <- PDM.CNV %>%
  summarise(across(everything(), ~ sum(. < -0.1 | . > 0.1)))

CNV.event.SCM <- SCM.CNV %>%
  summarise(across(everything(), ~ sum(. < -0.1 | . > 0.1)))

data.CNV <- data.CNV[,which(CNV.event.PDM>2 | CNV.event.SCM>2)]
```

select features associated with subtypes (Spearman corrolation, R^2 > 0.25)

```{r, message=FALSE, warning=FALSE}
data.CNV.cor <- cor(data.CNV, method = "spearman") %>%
  as.data.frame() %>%
  dplyr::filter(abs(Subtype) > 0.25) 

data.CNV.subset <- data.CNV %>%
  dplyr::select(rownames(data.CNV.cor)) 

# data.CNV.cor.PDM <- data.CNV.subtype.cor %>%
#   dplyr::filter(Subtype> 0.25)
```

correlation dendrogram of associated features

```{r, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
# spearman correlation on subset CN data
data.CNV.subset.cor <- cor(data.CNV.subset, method = "spearman")

# calculate distant matrix
dist_mat <- dist(data.CNV.subset.cor[-c(1), -c(1)], method = 'euclidean')

# hierarchical clustering 
hclust_avg <- hclust(dist_mat, method = 'average') %>%
  as.dendrogram

# dendgrogram data
dend_data <- label(dendro_data(hclust_avg, type = "rectangle")) %>%
  mutate(gene = str_split(label,pattern = "_", simplify = T)[,1],
         chromosome = str_split(label,pattern = "_", simplify = T)[,2])

chrom_seq <- paste("chr", c(1:22, "X"), sep = "")
chr_col <- colorRampPalette(c("grey20", "#E69F00", "#56B4E9", "tomato", "purple", "navyblue", "#CC79A7",  "#D55E00"))(23)


#pdf("figures/CNV_corplot_dendrogram.pdf", width = 8, height = 3)
ggplot(segment(dendro_data(hclust_avg, type = "rectangle"))) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) + 
  geom_text(data=dend_data, aes(label=label, x=x, y=-0.2, colour=chromosome), hjust = 1, angle = 90, size = 2.5, show.legend = F) +
  ylim(-4, 7) +
  scale_color_manual(values = chr_col) +
  theme_dendro()
#dev.off()
```

correlation plot of associated features

```{r, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
#pdf("CNV_corplot.pdf", width = 7.5, height = 7.5)
corrplot(data.CNV.subset.cor[c("Subtype", labels(hclust_avg)), c("Subtype", labels(hclust_avg))], tl.cex = 0.4, col=rev(COL2("PuOr", 200)), method = 'square', tl.col = "black")
#dev.off()
```  

PCA using associated features

```{r, fig.width=7.5, fig.height=7, message=FALSE, warning=FALSE}
cnvsubset.pca <- prcomp(data.CNV.subset[,-c(1)], scale. = T, center = T)
pca.res <- get_pca_var(cnvsubset.pca)

p1 <- fviz_eig(cnvsubset.pca)
p2 <- ggbiplot(cnvsubset.pca, var.axes = F, groups = factor(ifelse(data.CNV.subset$Subtype==1, "PDM", "metastasis"), labels = c("PDM", "metastasis")), choices = c(1,2), scale = 0)+
  scale_color_manual(values=c("#56B4E9", "#E69F00"))
p3 <- fviz_contrib(cnvsubset.pca, choice = "var", 
                   top =  sum(pca.res$contrib[,"Dim.1"] > 100/length(pca.res$contrib[,"Dim.1"])), axes = 1) 
toprow <- plot_grid(p1, p2, labels = c('A', 'B'), label_size = 12)

# pdf("figures/CNV_PCA_correlated_features.pdf", width = 7.5, height = 8)
plot_grid(toprow, p3, labels = c('', 'C'), label_size = 12, nrow = 2)
# dev.off()
```  


### cluster all CN features

hierarchical clustering of CN features

```{r, fig.width=15, fig.height=5, message=FALSE, warning=FALSE}
chrom_seq <- paste("chr", c(1:22, "X"), sep = "")
chr_col <- colorRampPalette(c("grey20", "#E69F00", "#56B4E9", "tomato", "purple", "navyblue", "#CC79A7",  "#D55E00"))(23)

data.CNV.t <- t(data.CNV[,-c(1)])
# data.CNV.cor <- cor(data.CNV[, -c(1)], method = "spearman") 

dist_mat <- dist(data.CNV.t, method = 'manhattan')
hclust_avg <- hclust(dist_mat, method = 'complete')
plot(hclust_avg)
```

Group the clustered CN features. Features are grouped by hierarchical clustering (threshold indicated as a red dashed line). Text/dot colours indicate the chromosomal number of the features.

```{r, fig.width=15, fig.height=5, message=FALSE, warning=FALSE}
dend_data <- label(dendro_data(hclust_avg, type = "rectangle")) %>%
  mutate(gene = str_split(label,pattern = "_", simplify = T)[,1],
         chromosome = str_split(label,pattern = "_", simplify = T)[,2], 
         chromosome = factor(chromosome, levels = chrom_seq)) 


#pdf("figures/CNV_dendrogram_all_features_text.pdf", width = 15, height = 5)
ggdendrogram(hclust_avg, labels = F) +
  geom_text(data=dend_data, aes(label=label, x=x, y=-0.2, colour=chromosome), hjust = 1, angle = 90, size = 2.5, show.legend = F) +
  geom_hline(yintercept=5.5, linetype="dashed", color = "red") +
  scale_colour_manual(values = chr_col) +
  theme_dendro()
#dev.off()

#pdf("figures/CNV_dendrogram_all_features_dot.pdf", width = 15, height = 5)
ggdendrogram(hclust_avg, labels = F) +
  geom_point(data=dend_data, aes(x=x, y=-0.2, colour=chromosome), size = 2, show.legend = T) +
  geom_hline(yintercept=5.5, linetype="dashed", color = "red") +
  scale_colour_manual(values = chr_col) +
  theme_dendro()
#dev.off()
```  

check the number of clusters, and number of unique chromosomes in each cluster

```{r, message=FALSE, warning=FALSE}
cut_tree <- cutree(hclust_avg, h = 5.5)
hclust.cluster = data.frame(feature = names(cut_tree), cluster = as.numeric(cut_tree))%>%
  mutate(gene = str_split(feature,pattern = "_", simplify = T)[,1],
         chromosome = str_split(feature,pattern = "_", simplify = T)[,2]) 

hclust.cluster.summary <- hclust.cluster %>%
  group_by(cluster) %>%
  summarise(unique_chr = n_distinct(chromosome),
            gene_number = n(), 
            chr = unique(chromosome)) %>%
  mutate(cluster = paste("cluster_", cluster, sep = "")) 

cat("Total cluster number: ", length(unique(hclust.cluster$cluster)), "\n")
cat("Max number of chromosome in each cluster: ", max(hclust.cluster.summary$unique_chr))
``` 

calculate average copy ratio of each cluster

```{r, message=FALSE, warning=FALSE}
data.CNV.t <- t(data.CNV[,-c(1)]) %>%
  as.data.frame() %>%
  mutate(feature = names(data.CNV[,-c(1)])) %>%
  left_join(hclust.cluster, by = "feature")

data.CNV.cluster <- data.CNV.t %>%
  group_by(cluster) %>%
  summarise_at(.vars = rownames(data.CNV), .funs = mean)

data.CNV.cluster.t <- t(data.CNV.cluster[,-c(1)]) %>%
  as.data.frame()
names(data.CNV.cluster.t) <- paste("cluster_", data.CNV.cluster$cluster, sep = "")
```  

write CN clusters

```{r}
data.CNV.cluster.t <- data.CNV.cluster.t %>%
  mutate(sample = rownames(data.CNV.cluster.t), .before = cluster_1)
#write.csv(data.CNV.cluster.t, file = "tables/CNV_clusters_avg_log2R.csv", row.names = F)
```











